Utilization of Reductions and Abstraction Elimination in Typed Genetic Programming

### 2014-04-15 tex/gecco14/gecco14.tex
cla

@@

\newcommand{\Lets}{Let us\xspace}
\newcommand{\lets}{let us\xspace}
\newcommand{\lterm}{$\lambda$-term\xspace}
\newcommand{\lterms}{$\lambda$-terms\xspace}
\newcommand{\lhead}{$\lambda$-head\xspace}
\newcommand{\lheads}{$\lambda$-heads\xspace}
\newcommand{\la}{\leftarrow\xspace}
\newcommand{\Lp}  {\Lambda^{\prime}\xspace}
\newcommand{\tur}[3]{#1\vdash{}#2 \colon #3}
\newcommand{\turst}[3]{$#1\vdash{}#2:#3$\xspace}
\newcommand{\GMS}{\turst{\Gamma}{M}{\sigma}}
\newcommand{\atTree}{@-tree\xspace}
\newcommand{\setDots}[2]{ \lbrace #1 , \dots , #2 \rbrace}
\newcommand{\lh}[1]{\lambda #1}
\newcommand{\sexprTree}{sexpr-tree\xspace}
\newcommand{\SexprTree}{Sexpr-tree\xspace}
\newcommand{\then}{\Rightarrow\xspace}
\newcommand{\lamb}[2]{( \lambda \, #1 \, . \, #2 )}
\newcommand{\lam}[2]{\lambda \, #1 \, . \, #2}
\newcommand{\ST}{\mathop{\mathrm{ST}}}
\newcommand{\FV}{\mathop{\mathrm{FV}}}
\newcommand{\Scomb }{\mathbf{S}}
\newcommand{\Kcomb }{\mathbf{K}}
\newcommand{\Icomb }{\mathbf{I}}
\newcommand{\bbarr}{\twoheadrightarrow_\beta}
\newcommand{\barr}{\rightarrow_\beta}
\newcommand{\beq}{=_\beta}
\newcommand{\eearr}{\twoheadrightarrow_\eta}
\newcommand{\earr}{\rightarrow_\eta}
\newcommand{\eeq}{=_\eta}
\newcommand{\bearr}{\rightarrow_{\beta\eta}}
\newcommand{\bbeearr}{\twoheadrightarrow_{\beta\eta}}
\newcommand{\beeq}{=_{\beta\eta}}
\newcommand{\etar}{\twoheadrightarrow_\eta}
\newcommand{\ered}{$\eta$-reduction\xspace}
\newcommand{\bnf}{$\beta$-\textit{nf}\xspace}
\newcommand{\enf}{$\eta$-\textit{nf}\xspace}
\newcommand{\eenf}{$\eta^{-1}$-\textit{nf}\xspace}
\newcommand{\beenf}{$\beta\eta^{-1}$-\textit{nf}\xspace}
\newcommand{\benf}{$\beta\eta$-\textit{nf}\xspace}
\newcommand{\bredex}{$\beta$-redex\xspace} 
\newcommand{\lnf}{\textit{lnf}\xspace}
\newcommand{\Ae}{\mathop{\mathrm{\AE}}}
\newcommand{\Bcomb }{\mathbf{B}}   
\newcommand{\BBcomb }{\mathbf{B*}}
\newcommand{\Ccomb }{\mathbf{C}}   
\newcommand{\CCcomb }{\mathbf{C'}}
\newcommand{\SScomb }{\mathbf{S'}}
\newcommand{\ar}{\rightarrow\xspace}
\newcommand{\T}{\mathbb{T}\xspace}
\newcommand{\C}{\mathbb{C}\xspace}
\newcommand{\Real}{\mathbb{R}}
\newcommand{\gar}{\longmapsto}
\newcommand{\bRedex}{$\beta$-redex\xspace}
\newcommand{\bRedexes}{$\beta$-redexes\xspace}
\newcommand{\bArrow}{\rightarrow_\beta\xspace}
\newcommand{\eArrow}{\rightarrow_\eta\xspace}
\newcommand{\eeArrow}{\rightarrow_{\eta^{-1}}\xspace}

@@

\conferenceinfo{GECCO'14,} {July 12-16, 2014, Vancouver, BC, Canada.}
\CopyrightYear{2014}
\crdata{TBA}
\clubpenalty=10000
\widowpenalty = 10000

@@

% \numberofauthors{2} 
%
%  \author{
%  % 1st. author
%  \alignauthor
%  Tom\'{a}\v{s} K\v{r}en\\
%    \affaddr{Faculty of Mathematics and Physics}\\
%    \affaddr{Charles University in Prague}\\
%    \affaddr{Malostransk\'e n\'am\v{e}st\'\i~25, 11000,}\\
%    \affaddr{Prague, Czech Republic}\\
%    \email{tomkren@gmail.com}
%  % 2nd. author
%  \alignauthor
%  Roman Neruda\\
%    \affaddr{Institute of Computer Science}\\
%    \affaddr{Academy of Sciences of the Czech Republic}\\
%    \affaddr{Pod Vod\'arenskou v\v{e}\v{z}\'\i~2, 18207,}\\
%    \affaddr{Prague, Czech Republic}\\
%    \email{roman@cs.cas.cz}
%  }

@@

\maketitle
\begin{abstract}
Lambda calculus representation of programs offers a more expressive alternative to traditional S-expressions. In this paper we discuss advantages of this representation coming from the use of reductions (beta and eta) and a way to overcome disadvantages caused by variables occurring in the programs by use of the abstraction elimination algorithm. We discuss the role of those reductions in the process of generating initial population and propose several crossover approaches, two of which are based on abstraction elimination and compared using the even parity benchmark problem. 
%including novel approach to crossover operator based both on reductions and abstraction elimination. 
%The design goal of this operator is to turn the disadvantage of abstraction elimination  - possibly quadratic increase of program size - into a virtue; our approach leads to more crossover points. At the same time, utilization of reductions provides offspring of small sizes.
\end{abstract}

@@

\keywords{genetic programming, lambda calculus}

@@

Genetic programming (GP) represents an efficient method for automatic generating of programs by means of evolutionary techniques~\cite{koza92,koza03}. Early attempts to enhance the GP approach with the concept of types include the seminal work~\cite{montana95} where the ideas from Ada programming language were used to define a so-called strongly typed GP.   
Usage of types naturally opens door to enriching S-expressions,
the traditional GP representation of individuals, with concepts from
lambda calculus, which is a simple yet powerful functional, mathematical and programming language extensively used in type theory. Such attempts have shown to be successful \cite{yu01}.

@@

In our system we use simply typed lambda calculus. Since abstraction elimination and reductions do not involve many type related issues it seems to us as a sufficient choice of type system. Techniques described here would be analogous for a more sophisticated type system and can be directly applied to such systems as the Hindley–Milner type system.

@@

%Pro inspiraci :
%\nehodi{The key issue in the lambda calculus approach to enrich GP with types is the method of individual generation. During the expansion phase the set of unfinished terms can be browsed with respect to various search strategies. Our approach to this problem aims to utilize the full arsenal given by the simply typed lambda calculus. Thus, the natural idea is to employ an exhaustive systematic search. On the other hand, if we were to mimic the standard GP approach, a quite arbitrary yet common and successful ramped half-and-half generating heuristic~\cite{fg} should probably be used. These two search methods in fact represent boundaries between which we will try to position our parameterized solution that allows us to take advantage of both strategies. This design goal also differentiate our approach from the three state of the art proposals for typed GP known to us that are discussed in the following section. Our proposed \emph{geometrical search strategy} described in this paper is such a successful hybrid mixture of random and systematic exhaustive search. Experiments show that it is also very efficient dealing with one of the traditional GP scarecrows - the bloat problem.}

The rest of the paper is organized as follows: The next section briefly discusses related work in the field of typed GP, while section~\ref{preliminaries} introduces necessary notions.
The body of the paper is presented in sections \ref{generating} and \ref{crossover}; the first discusses the term generating 
procedure and its relations
with reductions, the second describes several approaches to lambda term
crossover and how they solve the variable problem. Section~\ref{experiments} examines those ideas on the even parity benchmark problem, while the paper is concluded by section~\ref{conclusions}.

@@

Yu presents a GP system utilizing
polymorphic higher-order functions\footnote{Higher-order 
function is a function taking another function as 
input parameter.} and lambda abstractions  \cite{yu01}.
An important point of interest in the work is use of
\texttt{foldr} function as a tool for \textit{implicit recursion}, i.e. recursion without explicit recursive calls. The terminal set for constructing lambda abstraction subtrees  is limited to use only constants and variables of that particular lambda abstraction, i.e., outer variables are not allowed to be used as terminals in this work. This is a significant difference from our approach since we permit all well-typed normalized \lterms. This difference also causes the need for different crossover operators.

Briggs and O’Neill present a technique 
utilizing typed GP with combinators \cite{kes}.
The difference between approach presented in their work
and our approach is that they generate terms in a straightforward way directly from the library of combinators, without any use of lambda abstractions. They use the Hindley–Milner type system. They also present interesting concept of \textit{Generalized genetic operator} based on term generation. The combinator based genetic algorithm has the advantage of
avoiding the need of dealing with variables.

Binard and Felty use even stronger type system (\textit{System F}) \cite{binard2008genetic}. But with the increasing power of the type system also comes an increasing difficulty of term generation. For this reason, the evolution in this work takes interesting and nonstandard shape (fitness is associated with \textit{genes} which are evolved together with \textit{species} which together participate in creation of individuals). This differs from our approach, which tries to be a generalization of the standard GP\cite{koza92}.

In contrast with the above mentioned works, our approach uses very simple type system (simply typed lambda calculus) and concentrates on process of generation  
able to generate all possible well-typed normalized lambda terms. In order to do so we use technique based on \textit{inhabitation machines} described by Barendregt \cite{barendregt10}.

@@

@@

For better readability, 
$M_1~M_2~M_3~\dots~M_n$ will be an abbreviation for
$(\dots((M_1~M_2)~M_3)~\dots~M_n)$,
and $\lam{x_1 x_2 \dots x_n }{M}$ will be an abbreviation for 
$\lamb{x_1}{\lamb{x_2}{\dots\lamb{x_n}{M}\dots}}$.

@@

In order to perform computation, there must be some
mechanism for term evaluation. In $\lambda$-calculus there
is a \mbox{$\beta$-reduction} procedure for this reason.\\

@@

In other words, $\beta$-reduction is the process 
of insertion of arguments supplied to a function into 
its body. \\

@@

Relation $\bearr\;=\;\barr \cup \earr$. 
(Relation $R = \{\;(a,b)\;|\;a\;R\;b\;\}$.)
Similarly as for $\bbarr$ and $\beq$ we can define relations 
$\eearr$, $\eeq$,$\bbeearr$ and $\beeq$.\\

$\eta^{-1}$-reduction (also called $\eta$-expansion) is 
the reduction converse to $\eta$-reduction.
It is defined as follows.  
$$M \eeArrow \lamb{x}{(M~x)} \textbf{ ~~~~if } x \not\in FV(M) $$

@@

\begin{easylist}[enumerate]
& A \lterm is a \textit{$\beta$-normal form} (\bnf) 
if it does not have a $\beta$-redex as subterm.
& A \lterm M \textit{has} a \bnf if $M \beq N$
and $N$ is a \bnf.\\
\end{easylist}
\end{definition}
A normal form may be thought of as a result of a term evaluation. 
Normalization of the term $M$ is the process of finding normal form of $M$.

Similarly we can define \enf and \benf.

@@

\begin{definition}
A statement $M\colon\sigma$ is \textit{derivable from}
a context $\Gamma$ (notation 
\mbox{$\Gamma\vdash{}M\colon\sigma$}) 
if it can be produced by the following rules.
\begin{align*}
x : \sigma \in \Gamma &~\then~ \tur{\Gamma}{x}{\sigma}\\
\tur{\Gamma}{M}{\sigma \ar \tau}~,~\tur{\Gamma}{N}{\sigma} 
&~\then~ \tur{\Gamma}{(M~N)}{\tau}\\  
\tur{\Gamma,x:\sigma}{M}{\tau}
&~\then~ \tur{\Gamma}{\lamb{x}{M}}{\sigma \ar \tau} 
\end{align*}
\end{definition}

@@

\begin{proposition}
If $M$ has a \bnf, 
%which according to Theorem 2B.4 is always the case, 
then it also has a unique \lnf, 
which is also its unique \beenf.
\end{proposition}

@@

\subsection{Abstraction elimination}

\textit{Abstraction elimination} is a process of transforming 
an arbitrary \lterm into \lterm that contains no lambda abstractions
and no bound variables.
The newly produced \lterm may contain function applications, 
free symbols from former \lterm and some new symbols standing for 
combinators $\Scomb$, $\Kcomb$ and $\Icomb$. \\

@@

\Lets describe transformation $\Ae$ performing this 
process.
%\footnote{In our implementation we must also deal with types, because our individuals are annotated with types. But since this process is fairly straightforward but cumbersome to describe, we will skip explaining it here. Reader interested in seeing it can see the source code of our implementation. }
\begin{align*}
\Ae[x]           &= x &\\[0.4em]
\Ae[\,(M\,N)\,]  &= (\Ae[M]\;\;\Ae[N]) &\\[0.4em]
\Ae[\lam{x}{x}]  &= \Icomb &\\
\Ae[\lam{x}{M}]  &= (\Kcomb~\Ae[M]) &\textbf{if } x \not\in \FV(M)\\
\Ae[\lam{x}{\lamb{y}{M}}] &= \Ae[\lam{x}{\Ae[\lam{y}{M}]}]  
&\textbf{if } x \in \FV(M)\\
\Ae[\lam{x}{(M\,N)}] &= (\Scomb~\Ae[\lam{x}{M}]~\Ae[\lam{x}{N}])  
&\textbf{if } x \in \FV(M)\\
&&\vee~ x \in \FV(N)\\
\end{align*}

@@

The algorithm presented here is a simple version of this process. 
More optimized version (used in our \textit{hybrid} crossover described below), by means of the size of resulting term and its time performance is presented in \cite{jones87}.

\section{Term generating and reductions}
\label{generating}

This section is focused on the individual generating method producing
terms in their long normal form, which can be understood as a straight 
generalization of S-expressions into lambda calculus. 
We discuss the relation between terms in \lnf with beta and eta reductions,
the advantages of such representation and we also show how to easily 
transform such terms into short \benf.

@@

For every atomic type $\alpha$, context $\Gamma$ and types
$\tau_1 , \dots , \tau_n$ (where $n > 0$) 
we have a rule of the second form:\\

$
( \tau_1 \ar \dots \ar \tau_n \ar \alpha , \Gamma )  
 \\ ~~~~~~~~\gar
(~\lambda~x_1~\dots~x_n~.~
( \alpha ; \Gamma , x_1:\tau_1 , \dots , x_n:\tau_n  )~)
$~\\

@@

By generating \lterms in \textit{lnf} we avoid generating 
\lterms $M$,$N$ such that $M \not= N$, but $M =_{\beta\eta} N$.
In other words, we avoid generating two programs with different 
source codes, but performing the same computation.

Every \lterm $M$ such that \GMS has its unique \lnf $L$, 
for which $L =_{\beta\eta} M$.
Therefore, the computation performed by \lterm $M$ 
is not omitted, because it is the same computation
as the computation performed by \lterm $L$. \\

@@

By generating \lterms in \lnf we avoid generating 
\lterms $M$,$N$ such that $M \not= N$ and $M =_{\eta} N$; 
but generating in \bnf does not have such a property..\\

@@

For better clarity \lets show the \ered and \bredex using \atTree{}s 
(a \lterm representation derived from the definition of \lterms 
where @ stands for \textit{function application}).\\
\textit{\ered}: \Tree [.$\lh{x}$ [.@ $M$ $x$ ] ] 
$\eArrow$
$M$ 
~~~~~~~~~~~
\textit{\bredex}: \Tree [.@ [.$\lh{x}$ $M$ ] $N$ ] \\

@@

Since reduction to \enf always leads to an unique term, it is a function.
In previous proposition is shown that \ered of \lnf
leads to a term in \benf.\\

In order to show that a function is bijection it is sufficient to show that it is
both \textit{injection} and \textit{surjection}.\\

@@

So there must be $M_1,M_2$ in \lnf such that $M_1 \not= M_2$
and $N$ in \benf such that $M_1 \etar N$, $M_2 \etar N$.
Therefore $M_1 =_\eta M_2$, 
so $M_1 =_{\beta\eta^{-1}} M_2$.
This contradicts with $M_1,M_2$ being distinct \lnf{}s.\\

@@

The crossover operation in standard GP is performed 
by swapping randomly selected subtrees in each parent 
S-expression.
For typed lambda terms two difficulties arise: Types and variables.

@@

The following example illustrates the problem. \Lets have these two
parent trees with selected nodes in bold.\\

@@

~\\The swap of subtrees results in following trees:\\

@@

~\\The problem is that variable $x_2$ in second tree
is not bound by any $\lambda$-head and since
it is not element of $\Gamma$, the second tree is not well-typed \lterm.

Generally speaking, the variable problem is caused by the fact the local variable is not defined in the new place, or the variable is defined in the new place, but has
some a different type. Let us list some possible approaches to solving this problem.

@@

The first thing to do in a standard subtree swapping crossover is to select random node in the first parent.

@@

~\\

>>>>

\newfont{\mycrnotice}{ptmr8t at 7pt}
\newfont{\myconfname}{ptmri8t at 7pt}
\let\crnotice\mycrnotice%
\let\confname\myconfname%

\permission{Permission to make digital or hard copies of all or part of this work for personal or classroom use is granted without fee provided that copies are not made or distributed for profit or commercial advantage and that copies bear this notice and the full citation on the first page. Copyrights for components of this work owned by others than the author(s) must be honored. Abstracting with credit is permitted. To copy otherwise, or republish, to post on servers or to redistribute to lists, requires prior specific permission and/or a fee. Request permissions from permissions@acm.org.}
\conferenceinfo{GECCO'14,}{July 12--16, 2014, Vancouver, BC, Canada. \\
{\mycrnotice{Copyright is held by the owner/author(s). Publication rights licensed to ACM.}}}
\copyrightetc{ACM \the\acmcopyr}
\crdata{978-1-4503-2662-9/14/07\ ...\$15.00.\\
http://dx.doi.org/10.1145/2576768.2598361}

\clubpenalty=10000 
\widowpenalty = 10000

%\begin{document}

%.....

@@

\newcommand{\superscript}[1]{\ensuremath{^{\textrm{#1}}}}
\def\sharedaffiliation{\end{tabular}\newline\begin{tabular}{c}}

\def\wu{\superscript{*}}
\def\wg{\superscript{\dag}}

\newcommand{\Lets}{Let us\xspace}
\newcommand{\lets}{let us\xspace}
\newcommand{\lterm}{$\lambda$-term\xspace}
\newcommand{\lterms}{$\lambda$-terms\xspace}
\newcommand{\lhead}{$\lambda$-head\xspace}
\newcommand{\lheads}{$\lambda$-heads\xspace}
\newcommand{\la}{\leftarrow\xspace}
\newcommand{\Lp}  {\Lambda^{\prime}\xspace}
\newcommand{\tur}[3]{#1\vdash{}#2 \colon #3}
\newcommand{\turst}[3]{$#1\vdash{}#2:#3$\xspace}
\newcommand{\GMS}{\turst{\Gamma}{M}{\sigma}}
\newcommand{\atTree}{@-tree\xspace}
\newcommand{\setDots}[2]{ \lbrace #1 , \dots , #2 \rbrace}
\newcommand{\lh}[1]{\lambda #1}
\newcommand{\sexprTree}{sexpr-tree\xspace}
\newcommand{\SexprTree}{Sexpr-tree\xspace}
\newcommand{\then}{\Rightarrow\xspace}
\newcommand{\lamb}[2]{( \lambda \, #1 \, . \, #2 )}
\newcommand{\lam}[2]{\lambda \, #1 \, . \, #2}
\newcommand{\ST}{\mathop{\mathrm{ST}}}
\newcommand{\FV}{\mathop{\mathrm{FV}}}
\newcommand{\Scomb }{\mathbf{S}}
\newcommand{\Kcomb }{\mathbf{K}}
\newcommand{\Icomb }{\mathbf{I}}
\newcommand{\bbarr}{\twoheadrightarrow_\beta}
\newcommand{\barr}{\rightarrow_\beta}
\newcommand{\beq}{=_\beta}
\newcommand{\eearr}{\twoheadrightarrow_\eta}
\newcommand{\earr}{\rightarrow_\eta}
\newcommand{\eeq}{=_\eta}
\newcommand{\bearr}{\rightarrow_{\beta\eta}}
\newcommand{\bbeearr}{\twoheadrightarrow_{\beta\eta}}
\newcommand{\beeq}{=_{\beta\eta}}
\newcommand{\etar}{\twoheadrightarrow_\eta}
\newcommand{\ered}{$\eta$-reduction\xspace}
\newcommand{\bnf}{$\beta$-\textit{nf}\xspace}
\newcommand{\enf}{$\eta$-\textit{nf}\xspace}
\newcommand{\eenf}{$\eta^{-1}$-\textit{nf}\xspace}
\newcommand{\beenf}{$\beta\eta^{-1}$-\textit{nf}\xspace}
\newcommand{\benf}{$\beta\eta$-\textit{nf}\xspace}
\newcommand{\bredex}{$\beta$-redex\xspace} 
\newcommand{\lnf}{\textit{lnf}\xspace}
%\newcommand{\Ae}{\mathop{\mathrm{\AE}}}
\newcommand{\Ae}{\mathop{\mathrm{AE}}}
\newcommand{\Bcomb }{\mathbf{B}}   
\newcommand{\BBcomb }{\mathbf{B*}}
\newcommand{\Ccomb }{\mathbf{C}}   
\newcommand{\CCcomb }{\mathbf{C'}}
\newcommand{\SScomb }{\mathbf{S'}}
\newcommand{\ar}{\rightarrow\xspace}
\newcommand{\T}{\mathbb{T}\xspace}
\newcommand{\C}{\mathbb{C}\xspace}
\newcommand{\Real}{\mathbb{R}}
\newcommand{\gar}{\longmapsto}
\newcommand{\bRedex}{$\beta$-redex\xspace}
\newcommand{\bRedexes}{$\beta$-redexes\xspace}
\newcommand{\bArrow}{\rightarrow_\beta\xspace}
\newcommand{\eArrow}{\rightarrow_\eta\xspace}
\newcommand{\eeArrow}{\rightarrow_{\eta^{-1}}\xspace}

@@

\setlength{\pdfpagewidth}{8.5in}
\setlength{\pdfpageheight}{11in}

%\conferenceinfo{GECCO'14,} {July 12-16, 2014, Vancouver, BC, Canada.}
%\CopyrightYear{2014}
%\crdata{TBA}
%\clubpenalty=10000
%\widowpenalty = 10000

@@

 \numberofauthors{2} 
  \author{
  % 1st. author
  \alignauthor
  Tom\'{a}\v{s} K\v{r}en\wu\wg\\
    \email{tomkren@gmail.com}\\
    ~\\
    \affaddr{{\wu}Faculty of Mathematics and Physics}\\
    \affaddr{Charles University in Prague}\\
    \affaddr{Malostransk\'e n\'am\v{e}st\'\i~25, 11000,}\\
    \affaddr{Prague, Czech Republic}
  % 2nd. author
  \alignauthor
  Roman Neruda\wg\\
    \email{roman@cs.cas.cz}\\
    ~\\
    \affaddr{{\wg}Institute of Computer Science}\\
    \affaddr{Academy of Sciences of the Czech Republic}\\
    \affaddr{Pod Vod\'arenskou v\v{e}\v{z}\'\i~2, 18207,}\\
    \affaddr{Prague, Czech Republic}    
  }

%\numberofauthors{2}
%\author{
%  \alignauthor \ Tom\'{a}\v{s} K\v{r}en\wu\wg\\
%  \email{tomkren@gmail.com}
%  %
%  \alignauthor \ Roman Neruda\wu\\
%  \email{roman@cs.cas.cz}
%  %
%  \sharedaffiliation
%  \begin{tabular}{ccc}
%    \affaddr{{\wu}Institute of Computer Science{\ }}      & & \affaddr{{\wg}Faculty of Mathematics and Physics{\ }} \\
%    \affaddr{Academy of Sciences of the Czech Republic}   & & \affaddr{Charles University in Prague} \\
%    \affaddr{Pod Vod\'arenskou v\v{e}\v{z}\'\i~2, 18207,} & & \affaddr{Malostransk\'e n\'am\v{e}st\'\i~25, 11000,} \\
%    \affaddr{Prague, Czech Republic}                      & & \affaddr{Prague, Czech Republic} \\
%  \end{tabular}
%}

@@

\maketitle
\begin{abstract}
Lambda calculus representation of programs offers a more expressive alternative to traditional S-expressions. In this paper we discuss advantages of this representation coming from the use of reductions (beta and eta) and a way to overcome disadvantages caused by variables occurring in the programs by use of the abstraction elimination algorithm. We discuss the role of those reductions in the process of generating initial population and propose two novel crossover operations based on abstraction elimination capable of handling general form of typed lambda term while being a straight generalization of the standard crossover operation.  We compare their performances using the even parity benchmark problem. 
%\red{Přidat víc o kříženích tak aby davál insight to the project. Možná i o tom že je to zobecnení std GP.}
%including novel approach to crossover operator based both on reductions and abstraction elimination. 
%The design goal of this operator is to turn the disadvantage of abstraction elimination  - possibly quadratic increase of program size - into a virtue; our approach leads to more crossover points. At the same time, utilization of reductions provides offspring of small sizes.
\end{abstract}

@@

\keywords{genetic programming; lambda calculus}

@@

Genetic programming (GP) represents an efficient method for automatic generating of programs by means of evolutionary techniques~\cite{koza92,koza03}. Early attempts to enhance the GP approach with the concept of types include the seminal work~\cite{montana95} where the ideas from Ada programming language were used to define a so-called strongly typed GP.   
Usage of types naturally opens door to enriching S-expressions,
the traditional GP representation of individuals, with concepts from
lambda calculus, which is a simple yet powerful functional, mathematical and programming language extensively used in type theory~\cite{yu01}.

@@

In our system we use a simply typed lambda calculus. Since abstraction elimination and reductions do not involve many type related issues it seems to us as a sufficient choice of a type system. Techniques described here would be analogous for a more sophisticated type system and can be directly applied to such systems as the Hindley–Milner type system.

In this paper we present an approach to typed GP over lambda calculus capable of operating with the full array of notions given by a simply typed lambda calculus while being a direct generalization of standard GP~\cite{koza92}. We discuss the advantages of generating lambda terms in their long normal form and present two novel crossover operations for simply typed lambda terms. Compared to the approach to typed GP over lambda calculus  proposed by 
Yu~\cite{yu01}, it uses more general term representation which is not restricting the use of outer local variables in the body of an anonymous function. Compared to the purely combinator approach by Briggs and 
O’Neill~\cite{kes}, it has the advantage of reducing the size of the search space in the term generation phase.

@@

The rest of the paper is organized as follows: The next section briefly discusses related work in the field of typed GP, while section~\ref{preliminaries} introduces necessary notions.
The body of the paper is presented in sections \ref{generating} and \ref{crossover}; the former discusses the term generating 
procedure and its relations
with reductions, the latter describes several approaches to lambda term
crossover, and how they solve the variable problem. Section~\ref{experiments} examines those ideas on the even parity benchmark problem, while the paper is concluded by section~\ref{conclusions}.

\vfill

@@

Yu presents a GP system utilizing
polymorphic higher-order functions
%\footnote{Higher-order function is a function taking another function as input parameter.} 
and lambda abstractions~\cite{yu01}.
An important point of interest in the work is the use of
\texttt{foldr} function as a tool for \textit{implicit recursion}, i.e., recursion without explicit recursive calls. The terminal set for constructing lambda abstraction subtrees  is limited to use only constants and variables of that particular lambda abstraction, i.e., outer variables are not allowed to be used as terminals in this work. This is a significant difference from our approach since we permit all well-typed normalized \lterms. This difference also causes the need for different crossover operators.

Briggs and O’Neill present a technique 
utilizing typed GP with combinators \cite{kes}.
The difference between the approach presented in their work
and our approach is that they generate terms in a straightforward way directly from the library of combinators, without any use of lambda abstractions. They use the Hindley–Milner type system. They also present an interesting concept of \textit{Generalized genetic operator} based on term generation. The combinator based genetic algorithm has the advantage of avoiding the need of dealing with variables.

Binard and Felty use an even stronger type system, the (\textit{System F}) \cite{binard2008genetic}. But with the increasing power of the type system also comes an increasing difficulty of term generation. For this reason, the evolution in this work takes interesting and nonstandard shape (fitness is associated with \textit{genes} which are evolved together with \textit{species} which together participate in creation of individuals). This differs from our approach which tries to be a straightforward generalization of the standard GP\cite{koza92}.

In contrast to the above mentioned works, our approach uses a very simple type system (the simply typed lambda calculus) and concentrates on process of generation  
able to generate all possible well-typed normalized lambda terms. In order to do so we use technique based on \textit{inhabitation machines} described by Barendregt\cite{barendregt10}.

@@

%\red{Buď přidat neformální komentář a/nebo odebrat nadbytečný věci.}

@@

To ensure better readability, 
$M_1~M_2~M_3~\dots~M_n$ will be an abbreviation for 
$(\dots((M_1~M_2)~M_3)~\dots~M_n)$.
And $\lam{x_1 x_2 \dots x_n }{M}$  for
$\lamb{x_1}{\lamb{x_2}{\dots\lamb{x_n}{M}\dots}}$.

@@

In order to perform computation, there must be some
mechanism for term evaluation. In $\lambda$-calculus there
is a \mbox{$\beta$-reduction} procedure for this reason.

@@

In other words, $\beta$-reduction is the process 
of insertion of arguments supplied to a function into 
its body.

@@

Relation\footnote{Relation $R = \{\;(a,b)\;|\;a\;R\;b\;\}$.} $\bearr\;=\;\barr \cup \earr$.
Similarly as for $\bbarr$ and $\beq$ we can define relations 
$\eearr$, $\eeq$,$\bbeearr$ and $\beeq$.

The $\eta^{-1}$-reduction (also called $\eta$-expansion) is 
the reduction converse to $\eta$-reduction.
It is defined as follows.  
$$M \eeArrow \lamb{x}{(M~x)} \textbf{ ~~~~if } x \not\in FV(M) $$

@@

\begin{easylist}[enumerate]
& A \lterm is a \textit{$\beta$-normal form} (\bnf) 
if it does not have a $\beta$-redex as subterm.
& A \lterm M \textit{has} a \bnf if $M \beq N$
and $N$ is a \bnf.
\end{easylist}
\end{definition}

A normal form may be thought of as a result of a term evaluation. 
Normalization of the term $M$ is the process of finding normal form of $M$. 
Similarly we can define \enf and \benf.

@@

\begin{definition}
A statement $M\colon\sigma$ is \textit{derivable from}
context $\Gamma$ (notation 
\mbox{$\Gamma\vdash{}M\colon\sigma$}) 
if it can be produced by the following rules.
\begin{align*}
x : \sigma \in \Gamma &~\then~ \tur{\Gamma}{x}{\sigma}\\
\tur{\Gamma}{M}{\sigma \ar \tau}~,~\tur{\Gamma}{N}{\sigma} 
&~\then~ \tur{\Gamma}{(M~N)}{\tau}\\  
\tur{\Gamma,x:\sigma}{M}{\tau}
&~\then~ \tur{\Gamma}{\lamb{x}{M}}{\sigma \ar \tau} 
\end{align*}
\end{definition}

@@

\begin{proposition}
If $M$ has a \bnf, 
%which according to Theorem 2B.4 is always the case, 
then it also has an unique \lnf, 
which is also its unique \beenf.
\end{proposition}

@@

\subsection{Abstraction elimination}
\textit{Abstraction elimination} is a process of transforming 
an arbitrary \lterm into \lterm that contains no lambda abstractions
and no bound variables.
The newly produced \lterm may contain function applications, 
free symbols from former \lterm and some new symbols standing for 
combinators $\Scomb$, $\Kcomb$ and $\Icomb$.

@@

\Lets describe transformation $\Ae$ performing this 
process.
%\footnote{In our implementation we must also deal with types, because our individuals are annotated with types. But since this process is fairly straightforward but cumbersome to describe, we will skip explaining it here. Reader interested in seeing it can see the source code of our implementation. }
\begin{align*}
\Ae[x]           &= x\\[0.4em]
\Ae[\,(M\,N)\,]  &= (\Ae[M]\;\;\Ae[N])\\[0.4em]
\Ae[\lam{x}{x}]  &= \Icomb\\
\Ae[\lam{x}{M}]  &= (\Kcomb~\Ae[M]) \textbf{ if } x \not\in \FV(M)\\
\Ae[\lam{x}{\lamb{y}{M}}] &= \Ae[\lam{x}{\Ae[\lam{y}{M}]}]  \textbf{ if } x \in \FV(M)\\
\Ae[\lam{x}{(M\,N)}] &= (\Scomb~\Ae[\lam{x}{M}]~\Ae[\lam{x}{N}])\\
&\textbf{ if } x \in \FV(M) \vee~ x \in \FV(N)\\
\end{align*}

@@

The algorithm presented here is a simple version of this process. 
More optimized version (used in our \textit{hybrid} cross\-over described below), by means of the size of resulting term and its time performance is presented in \cite{jones87}.

\section{Generating and reductions}
\label{generating}

This section is focused on the individual generating meth\-od producing
terms in their long normal form, which can be understood as a straight 
generalization of S-expressions into lambda calculus. 
We discuss the relation between terms in \lnf with beta and eta reductions,
the advantages of such representation and we also show how to easily 
transform such terms into short \benf.

@@

For every atomic type $\alpha$, context $\Gamma$ and types
$\tau_1 , \dots , \tau_n$ (where $n > 0$) 
we have a rule of the second form:

$
( \tau_1 \ar \dots \ar \tau_n \ar \alpha , \Gamma )  
 \\ ~~~~~~~~\gar
(~\lambda~x_1~\dots~x_n~.~
( \alpha ; \Gamma , x_1:\tau_1 , \dots , x_n:\tau_n  )~)
$

@@

By generating \lterms in \textit{lnf} we avoid generating 
\lterms $M$,$N$ such that $M \not= N$, but $M =_{\beta\eta} N$.
This means that $M$,$N$ correspond to two programs with different source codes, but both perform the same computation.

Every \lterm $M$ such that \GMS has its unique \lnf $L$, 
for which $L =_{\beta\eta} M$.
Therefore, the computation performed by \lterm $M$ 
is not omitted, because it is the same computation
as the computation performed by \lterm $L$.

@@

By generating \lterms in \lnf we avoid generating 
\lterms $M$,$N$ such that $M \not= N$ and $M =_{\eta} N$; 
but generating in \bnf does not have such a property.

@@

For better clarity \lets show the \ered and \bredex using \atTree{}s 
(a \lterm representation derived from the definition of \lterms 
where @ stands for \textit{function application}).

\textit{\ered}: \Tree [.$\lh{x}$ [.@ $M$ $x$ ] ] 
$\eArrow$
$M$ 
~~~~~~~~~~~
\textit{\bredex}: \Tree [.@ [.$\lh{x}$ $M$ ] $N$ ]

@@

Since reduction to \enf always leads to an unique term, it is a function.
In previous proposition is shown that \ered of \lnf
leads to a term in \benf.

In order to show that a function is bijection it is sufficient to show that it is
both \textit{injection} and \textit{surjection}.

@@

So there must be $M_1,M_2$ in \lnf such that $M_1 \not= M_2$
and $N$ in \benf such that $M_1 \etar N$, $M_2 \etar N$.
Therefore $M_1 =_\eta M_2$, 
so $M_1 =_{\beta\eta^{-1}} M_2$.
This contradicts with $M_1,M_2$ being distinct \lnf{}s.

@@

The crossover operation in standard GP is performed 
by swapping randomly selected subtrees in each parent 
S-expre\-ssion.
For typed lambda terms two difficulties arise: Types and variables.

@@

The following example illustrates the problem. \Lets have these two
parent trees with selected nodes in bold.

\vfill

@@

\vfill

The swap of subtrees results in following trees:

\vfill

@@

\vfill

The problem is that variable $x_2$ in second tree
is not bound by any $\lambda$-head and since
it is not element of $\Gamma$, the second tree is not well-typed \lterm.

Generally speaking, the variable problem is caused by the fact that the local variable is not defined in the new place, or the variable is defined in the new place, but has a different type. Let us list some possible approaches to solving this problem.

@@

The first thing to do in a standard subtree swapping cross\-over is to select random node in the first parent.

@@
